Counting And Confusion: Bayesian Rate Estimation With Multiple Populations

### 2014-10-01 response.txt

Dear Editor:

This is a response to the referee report on our manuscript DB11163 by Farr et al.

We apologize for the very delayed response.  We are grateful to the referee for a thorough reading of the manuscript and a number of useful comments.  We have incorporated the suggestions of the referee.  The text in bold represents the changes that have been made relative to the initial submission.  We also describe specific changes inline below.

>> ----------------------------------------------------------------------
>> Report of the Referee -- DB11163/Farr
>> ----------------------------------------------------------------------
>>
>> Referee report for manuscript DB11163 Farr
>> ==========================================
>>
>> The manuscript presents an application of Bayesian inference to the
>> problem of estimating the number of events in observed data which
>> belong to two independent populations, an example of a mixture model
>> with latent variables indicating the class of each event. It
>> illustrates the approach with examples from gravitational wave
>> searches and from fitting parameters of a stellar cluster superimposed
>> on a background. Although the authors only use simulated data in their
>> examples, the method employed is interesting and makes an advance on
>> the earlier methods used in the literature for approaching these
>> problems, although I do not think the statistics are novel in
>> themselves.
>>
>> The quality of writing and clarity of explanations is consistently
>> very high, and I believe the article is appropriate for publication in
>> PRD, but I have a few comments noted below that I recommend the
>> authors take into account.
>>
>> 1. There is a little confusion between number and rate in places. The
>> authors estimate rates of Poisson distributions throughout, i.e. the
>> number per unit time or area. The implicit factor of time or area
>> should be noted in the definition of R_f and R_b to avoid confusion,
>> and for dimensional correctness. They also occasionally refer to these
>> as numbers not rates in the text, e.g. after eq 34.

Whatever the units of f and \hat{f} [which depends on the space in which the corresponding rate density is measured], R_f provides the overall scaling amplitude in Eq. 9 and is therefore dimensionless (and similarly for R_b).  We have added a comment to this effect in the text.

>>
>> 2. Physical interpretations. I had difficulty identifying the key
>> result in the paper. The conclusions state the superiority of their
>> "framework", but if I wanted to use the framework where would I start?
>> Equation 18 gives the full posterior for the unknown variables so
>> perhaps that is the place. Or is it eqn 21? I suggest that the authors
>> emphasise the key results of their work with a view to how it might be
>> applied by the reader for his/her own particular problem.

Indeed, (18) and (21) are the two key equations; which one is used depends on whether the marginalized probability distributions associated with individual events are of interest.  We have made this clear in both the introduction and the conclusion.

>
>> 3. Using $f_i$ for the flag variables seems unnecessary when
>> $\hat{f}(x_i)$ and $\tilde{f}$ are also used. Can you choose a
>> different symbol?

We have replaced $f_i$ with $g_i$.

>>
>> 4. Before eqn 32 the start of the paragraph repeats some of the end of
>> the previous paragraph, suggest rewording.
>>
>> 5. I didn't understand the comment about comparing rates for different
>> thresholds just before eqn 35. Could you expand on this paragraph,
>> perhaps with an example?
>>
>> 6. Below eqn 37, "...average of the Bayesian estimate $R_B$" should be
>> an estimatOR. Likewise in the next sentence all threshold values will
>> yield, on average, the same POINT estimate.
>>
>> 7. After eqn 59 the method of time slides is mentioned, and it is
>> stated the background in time slides will be representative of the
>> true background. I don't think this is true when there are real events
>> in the unslid data which can be slid into coincidence with background
>> triggers. This is hardly a key point of the paper but it might be
>> worth mentioning that here.
>>
>> 8. The x-axis in figure 2 is very unclear, having only 1 reference
>> point. I suggest adding another.
>>
>> 9. The axes for figure 9 are illegibly small.
>>
>> Section V.B: As this is the section of most relevance for
>> gravitational wave analyses, it could benefit from some expansion, or
>> perhaps a separate paper if it grows too large. I would also like the
>> authors to clarify these questions below.
>>
>> 10. The authors do not mention the effect of overlapping template
>> banks on the foreground. Won't there also be correlated foreground
>> events as well as background?
>>
>> 11. It is stated that Neff is a property of the background
>> distribution, but shouldn't it be a property of the template bank
>> itself, or a property of the model that is being analysed?
>>
>> 12. It is stated that they used the same dataset as sec. V.A, with the
>> same number of events. Doesn't this mean that there were actually no
>> correlated events in the distributions? A more convincing example
>> would actually simulate the new background.
>>

———————————————————————————————————
Report of the Referee -- DB11163/Farr
----------------------------------------------------------------------

Referee report for manuscript DB11163 Farr
==========================================

The manuscript presents an application of Bayesian inference to the
problem of estimating the number of events in observed data which
belong to two independent populations, an example of a mixture model
with latent variables indicating the class of each event. It
illustrates the approach with examples from gravitational wave
searches and from fitting parameters of a stellar cluster superimposed
on a background. Although the authors only use simulated data in their
examples, the method employed is interesting and makes an advance on
the earlier methods used in the literature for approaching these
problems, although I do not think the statistics are novel in
themselves.

The quality of writing and clarity of explanations is consistently
very high, and I believe the article is appropriate for publication in
PRD, but I have a few comments noted below that I recommend the
authors take into account.

1. There is a little confusion between number and rate in places. The
authors estimate rates of Poisson distributions throughout, i.e. the
number per unit time or area. The implicit factor of time or area
should be noted in the definition of R_f and R_b to avoid confusion,
and for dimensional correctness. They also occasionally refer to these
as numbers not rates in the text, e.g. after eq 34.

2. Physical interpretations. I had difficulty identifying the key
result in the paper. The conclusions state the superiority of their
"framework", but if I wanted to use the framework where would I start?
Equation 18 gives the full posterior for the unknown variables so
perhaps that is the place. Or is it eqn 21? I suggest that the authors
emphasise the key results of their work with a view to how it might be
applied by the reader for his/her own particular problem.

3. Using $f_i$ for the flag variables seems unnecessary when
$\hat{f}(x_i)$ and $\tilde{f}$ are also used. Can you choose a
different symbol?

4. Before eqn 32 the start of the paragraph repeats some of the end of
the previous paragraph, suggest rewording.

5. I didn't understand the comment about comparing rates for different
thresholds just before eqn 35. Could you expand on this paragraph,
perhaps with an example?

6. Below eqn 37, "...average of the Bayesian estimate $R_B$" should be
an estimatOR. Likewise in the next sentence all threshold values will
yield, on average, the same POINT estimate.

7. After eqn 59 the method of time slides is mentioned, and it is
stated the background in time slides will be representative of the
true background. I don't think this is true when there are real events
in the unslid data which can be slid into coincidence with background
triggers. This is hardly a key point of the paper but it might be
worth mentioning that here.

8. The x-axis in figure 2 is very unclear, having only 1 reference
point. I suggest adding another.

9. The axes for figure 9 are illegibly small.

Section V.B: As this is the section of most relevance for
gravitational wave analyses, it could benefit from some expansion, or
perhaps a separate paper if it grows too large. I would also like the
authors to clarify these questions below.

10. The authors do not mention the effect of overlapping template
banks on the foreground. Won't there also be correlated foreground
events as well as background?

11. It is stated that Neff is a property of the background
distribution, but shouldn't it be a property of the template bank
itself, or a property of the model that is being analysed?

12. It is stated that they used the same dataset as sec. V.A, with the
same number of events. Doesn't this mean that there were actually no
correlated events in the distributions? A more convincing example
would actually simulate the new background.

### 2014-10-01 many.tex
Finished edits in response to referee's comments.

\author{Will M. Farr} \email{w-farr@northwestern.edu}
\homepage{http://faculty.wcas.northwestern.edu/will-farr/}
\affiliation{Center for Interdisciplinary Exploration and Research in
  Astrophysics\\ Department of Physics and Astronomy\\ Northwestern
  University, 2145 Sheridan Road, Evanston, IL 60208}

@@

The task of estimating rates of events when a mixture of foreground
and background events is present in data is a common one in physical
and astrophysical applications.  This problem comes up, among others,
in gravitational-wave data analysis
\cite[e.g.,][]{s5-lowmass,s5-highmass,s5-highmass-erratum,s5-lowmass-2,s6-lowmass,s6-highmass}
and in astronomical observations of a field of objects of mixed
provenance \cite{DeGennaro2009}.  In this paper, we introduce a robust
formalism for estimating event rates from the data when the shape of
foreground and background distributions are known (or parameterized),
but the provenance of individual events as either background or
foreground is unknown.

@@

@@

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process with respective differential
rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent.
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(n = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligible
  \begin{equation}
    P(n = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f \equiv
F(\infty,\theta)$ and $R_b \equiv B(\infty,\theta)$ are the total
number of foreground and background events expected and $\hat{f}(x,
\theta')$ and $\hat{b}(x, \theta')$ are the likelihood of obtaining an
event with ranking statistic $x$ under the foreground and background
distributions.  In what follows, we will drop the prime, using
$\theta$ to denote all parameters of the rate distributions except
$R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 (background) or 1 (foreground).  These ``state'' flags are
parameters in our model, along with $R_f$, $R_b$, and $\theta$.  We
can marginalize over our uncertainty in the state of any given event
by summing posteriors over $f_i = \mathset{0,1}$.

Assuming time-ordered data, $d_{\rm to}$, in the following, Bayes'
theorem relates the posterior probability of the state flags, rates,
and shape parameters, $p\left(\mathset{f_i}, R_f, R_b, \theta | d_{\rm
  to}, N \right)$, the likelihood of the data, $p\left( d_{\rm to} |
\mathset{f_i}, N, R_f, R_b, \theta\right)$, and the prior probability
of state flags, rates and shape parameters before any data are
obtained, $p\left( \mathset{f_i}, N, R_f, R_b, \theta\right)$:
\begin{multline}
  \label{eq:Bayes}
  p\left(\mathset{f_i}, R_f, R_b, \theta | d_{\rm to}, N \right) \\ =
  \frac{p\left( d_{\rm to} | \mathset{f_i}, N, R_f, R_b, \theta\right)
    p\left(\mathset{f_i}, N, R_f, R_b, \theta\right)}{p(d_{\rm to},
    N)}.
\end{multline}
The normalization constant, called the evidence, $p(d_{\rm to}, N)$,
is independent of the state flags, rates, and shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is
\begin{multline}
  p\left( d_{\rm to} | \mathset{f_i}, N, R_f, R_b, \theta\right) \\ =
  \left[ \prod_{\mathset{i | f_i = 1}} \hat{f}\left(x_i, \theta\right)
    \right] \left[ \prod_{\mathset{i | f_i = 0}} \hat{b}\left( x_i,
    \theta\right) \right].
\end{multline}
This is the probability that the first observed event is a
fore/background event (if $f_1=1,0$) with ranking statistic $x_1$ and
the second observed event is a fore/background event (if $f_2=1,0$)
with ranking statistic $x_2$, etc. If the events are ordered by ranking
statistic the corresponding expression is more complicated, since
$x_1$ is now the event from foreground or background with the smallest
ranking statistic, etc. We will return to the statistic-ordered case
later.

The prior distribution can be factorized as
\begin{multline}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{f_i}, N, R_f, R_b, \theta\right) \\ = p\left(
  \mathset{f_i} | N, R_f, R_b\right) p\left(N |R_f, R_b\right)
  p\left(R_f, R_b, \theta\right) \\ = p\left( \mathset{f_i},N | R_f,
  R_b\right) p\left(R_f, R_b, \theta\right).
 \end{multline}

The probability that the $i$'th state flag is $f_i=1$ is given by
$R_f/(R_f+R_b)$, while the probability that it is zero is
$R_b/(R_f+R_b)$, provided the data are time-ordered as we have
assumed.  Then
\begin{multline}
p\left( \mathset{f_i} | N, R_f, R_b\right) \\ = \prod_{\mathset{i|f_i=1}}
\left(\frac{R_f}{R_f+R_b}\right) \prod_{\mathset{i|f_i=0}}
\left(\frac{R_b}{R_f+R_b}\right) \\ =
\left(\frac{R_f}{R_f+R_b}\right)^{N_f}
\left(\frac{R_b}{R_f+R_b}\right)^{N_b},
\end{multline}
where $N_f$ and $N_b$ are the numbers of foreground and background
flags, $N_f+N_b=N$.  Meanwhile,
\begin{equation}
p\left(N |R_f, R_b\right) = \frac{\left(R_f+R_b\right)^N}{N!}
e^{-(R_f+R_b)},
\end{equation}
since the distribution of total event number is a Poisson process with
rate $R_f+R_b$.  Combining these yields the conditional probability of
the flags on the rates:
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{f_i},N | R_f, R_b\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N!} \exp\left[ - \left(R_f + R_b\right) \right].
\end{equation}

@@

Putting everything together, the posterior is
\begin{multline}
  \label{eq:posterior}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d_{\rm to},N \right) =
  \frac{\alpha}{p(d_{\rm to},N)\,N!} \\ \times \left[ \prod_{\mathset{i | f_i =
        1}} R_f \hat{f}\left(x_i, \theta\right) \right] \left[
    \prod_{\mathset{i | f_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
    \right] \\ \times \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
When sampling the posterior, the first term, which is independent of
the parameters of interest, can be omitted and the equals sign
replaced by proportionality; however, we have kept this term
explicitly so that we can see the equivalence to ranking-statistic
ordered data. Once data have been observed, there is a unique loudness
ordering and time ordering of those events, and so there is a one to
one correspondence between a time-ordered posterior $p\left(
\mathset{f_i}, R_f, R_b, \theta | d_{\rm to},N \right)$ and the
corresponding statistic-ordered posterior $ p\left( \mathset{f_i},
R_f, R_b, \theta | d,N \right)$, which means $p\left( \mathset{f_i},
R_f, R_b, \theta | d,N \right) = p\left( \mathset{f_i}, R_f, R_b,
\theta | d_{\rm to},N \right)$. However, the evidence $p(d,N) =
N!\,p(d_{\rm to},N)$, since there are $N!$ ways in which $N$ events
with a given set of ranking statistics can be ordered in time.

The ranking-statistic ordered posterior can be computed directly by
assuming that the flags, $\mathset{f_i}$, are un-observed data and
treating the sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i =
  0}$ as samples from an inhomogeneous Poisson process.  For an
inhomogeneous Poisson process with rate function $r(y)$ (cumulative
rate $R(y)$), the likelihood of a set of samples $\mathset{y_i}$ is
given by
\begin{multline}
  p\left( \mathset{y_i} | r \right) \,{\rm d}^Ny_i = P\left(
  \textnormal{zero events below } y_1 \right) \\ \times P\left(
  \textnormal{one event between } y_1 \textnormal{ and } y_1 + {\rm
    d}y_1 \right) \\ \times P\left( \textnormal{zero events between }
  y_1+{\rm d}y_1 \textnormal{ and } y_2 \right) \ldots,
\end{multline}
so 
\begin{multline}
p\left(\mathset{y_i} | r \right) = \lim_{\delta y_i \to 0} \exp\left[ -
    R\left(y_1\right) \right] \left[ r\left( y_1 \right) +
    \order{\delta y_1} \right] \\ \times \exp\left[ - \left[ R\left( y_2
      \right) - R\left( y_1 + \delta y_1\right) \right] \right] \times
  \ldots \\ = \left[\prod_i r\left( y_i \right)\right] \exp\left[ -
    R\left( \infty \right) \right].
\end{multline}
Applying this once to the foreground samples, once to the background
samples and taking the product, we obtain $p(d,\mathset{f_i}, N | R_f,
R_b, \theta)$ and thus $p(\mathset{f_i}, R_f, R_b, \theta | d,N) =
p(d,\mathset{f_i}, N | R_f, R_b, \theta)\ p(R_f, R_b, \theta)\, /\,
p(d,N)$. With the identification $p(d,N) = N!\,p(d_{\rm to},N)$, as
justified above, we reproduce Eq.~\eqref{eq:posterior}.

We can marginalize the posterior over the flags, $f_i$, obtaining
\begin{multline}
  \label{eq:rate-shape-posterior}
  p\left( R_f, R_b, \theta | d, N \right) = \sum_{\mathset{f_i} \in
    \mathset{0,1}^N} p\left( \mathset{f_i}, R_f, R_b, \theta | d, N
  \right) \\ \propto \prod_{i} \left[ R_f \hat{f}\left(x_i, \theta\right)
    + R_b \hat{b}\left( x_i, \theta\right) \right] \\ \times \exp\left[-\left(
    R_f + R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
This expression is useful if we are only interested in rates and not
the probability that any particular event is foreground or background.
Unlike the full posterior (Eq.~\eqref{eq:posterior}),
Eq.~\eqref{eq:rate-shape-posterior} contains only continuous
parameters. We note that the terms that depend on the overall rate
parameters, $R_b$ or $R_f$, are of the form $R_b^{n-1/2} \exp(-R_b)$
and so marginalization over either $R_b$ or $R_f$ can be achieved
analytically using
\begin{equation}
I_n = \int_0^{\infty} x^{n-\frac{1}{2}} \, {\rm e}^{-x} {\rm d} x =
\frac{(2n-1)!!}{2^n} \sqrt{\pi}
\end{equation}
using the usual notation $(2n-1)!! \equiv (2n-1)(2n-3)\cdots1$.

@@

If $\tilde{b}(x_N,\theta) \ll \tilde{f}(x_N,\theta)$, we obtain the
result $(1- \hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be
understood as the statement that the rate of foreground events with
ranking statistic greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$,
is of order $1$, as expected. However, $\tilde{b}(x_N,\theta) = -{\rm
  d} [\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term may be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) \gg \tilde{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavored.  As we will see, the problem can also be avoided
in the context of the loudest-event framework by even very weak prior
information on the background rate, $R_b$, of the kind present in
nearly all experiments.

>>>>

\author{Will M. Farr} \email{wfarr@star.sr.bham.ac.uk}
\homepage{http://faculty.wcas.northwestern.edu/will-farr/}
\affiliation{Center for Interdisciplinary Exploration and Research in
  Astrophysics\\ Department of Physics and Astronomy\\ Northwestern
  University, 2145 Sheridan Road, Evanston, IL 60208}
\affiliation{School of
  Physics and Astronomy\\University of Birmingham\\Edgbaston B15 2TT
  Birmingham\\United Kingdom}

@@

The task of estimating rates of events when a mixture of foreground and background
events is present in data is a common one in physical and
astrophysical applications.  This problem comes up, among others, in
gravitational-wave data analysis
\cite[e.g.,][]{s5-lowmass,s5-highmass,s5-highmass-erratum,s5-lowmass-2,s6-lowmass,s6-highmass}
and in astronomical observations of a field of objects of mixed
provenance \cite{DeGennaro2009,Koposov2010}.  In this paper, we introduce a robust
formalism for estimating event rates from the data when the shape of
foreground and background distributions are known (or parameterized),
but the provenance of individual events as either background or
foreground is unknown.

@@

{\bf Our key results appear in Eq.~(\ref{eq:posterior}), which provides the joint posterior probability distribution on the foreground and background rates and shape parameters and the provenance of individual events as either foreground or background.  Eq.~(\ref{eq:rate-shape-posterior}) is a marginalized version of Eq.~(\ref{eq:posterior}), useful when the provenance of individual events is not relevant.  In practice, these posteriors are best sampled with stochastic techniques such as Markov chain Monte Carlo.}

@@

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process with respective differential
rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent.
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(n = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligible
  \begin{equation}
    P(n = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall {\bf dimensionless} 
amplitude of the foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f \equiv
F(\infty,\theta)$ and $R_b \equiv B(\infty,\theta)$ are the total
number of foreground and background events expected and $\hat{f}(x,
\theta')$ and $\hat{b}(x, \theta')$ are the likelihood of obtaining an
event with ranking statistic $x$ under the foreground and background
distributions.  In what follows, we will drop the prime, using
$\theta$ to denote all parameters of the rate distributions except
$R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $g_i$, which is
either 0 (background) or 1 (foreground).  These ``state'' flags are
parameters in our model, along with $R_f$, $R_b$, and $\theta$.  We
can marginalize over our uncertainty in the state of any given event
by summing posteriors over $g_i = \mathset{0,1}$.

Assuming time-ordered data, $d_{\rm to}$, in the following, Bayes'
theorem relates the posterior probability of the state flags, rates,
and shape parameters, $p\left(\mathset{g_i}, R_f, R_b, \theta | d_{\rm
  to}, N \right)$, the likelihood of the data, $p\left( d_{\rm to} |
\mathset{g_i}, N, R_f, R_b, \theta\right)$, and the prior probability
of state flags, rates and shape parameters before any data are
obtained, $p\left( \mathset{g_i}, N, R_f, R_b, \theta\right)$:
\begin{multline}
  \label{eq:Bayes}
  p\left(\mathset{g_i}, R_f, R_b, \theta | d_{\rm to}, N \right) \\ =
  \frac{p\left( d_{\rm to} | \mathset{g_i}, N, R_f, R_b, \theta\right)
    p\left(\mathset{g_i}, N, R_f, R_b, \theta\right)}{p(d_{\rm to},
    N)}.
\end{multline}
The normalization constant, called the evidence, $p(d_{\rm to}, N)$,
is independent of the state flags, rates, and shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is
\begin{multline}
  p\left( d_{\rm to} | \mathset{g_i}, N, R_f, R_b, \theta\right) \\ =
  \left[ \prod_{\mathset{i | g_i = 1}} \hat{f}\left(x_i, \theta\right)
    \right] \left[ \prod_{\mathset{i | g_i = 0}} \hat{b}\left( x_i,
    \theta\right) \right].
\end{multline}
This is the probability that the first observed event is a
fore/background event (if $g_1=1,0$) with ranking statistic $x_1$ and
the second observed event is a fore/background event (if $g_2=1,0$)
with ranking statistic $x_2$, etc. If the events are ordered by ranking
statistic the corresponding expression is more complicated, since
$x_1$ is now the event from foreground or background with the smallest
ranking statistic, etc. We will return to the statistic-ordered case
later.

The prior distribution can be factorized as
\begin{multline}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{g_i}, N, R_f, R_b, \theta\right) \\ = p\left(
  \mathset{g_i} | N, R_f, R_b\right) p\left(N |R_f, R_b\right)
  p\left(R_f, R_b, \theta\right) \\ = p\left( \mathset{g_i},N | R_f,
  R_b\right) p\left(R_f, R_b, \theta\right).
 \end{multline}

The probability that the $i$'th state flag is $g_i=1$ is given by
$R_f/(R_f+R_b)$, while the probability that it is zero is
$R_b/(R_f+R_b)$, provided the data are time-ordered as we have
assumed.  Then
\begin{multline}
p\left( \mathset{g_i} | N, R_f, R_b\right) \\ = \prod_{\mathset{i|g_i=1}}
\left(\frac{R_f}{R_f+R_b}\right) \prod_{\mathset{i|g_i=0}}
\left(\frac{R_b}{R_f+R_b}\right) \\ =
\left(\frac{R_f}{R_f+R_b}\right)^{N_f}
\left(\frac{R_b}{R_f+R_b}\right)^{N_b},
\end{multline}
where $N_f$ and $N_b$ are the numbers of foreground and background
flags, $N_f+N_b=N$.  Meanwhile,
\begin{equation}
p\left(N |R_f, R_b\right) = \frac{\left(R_f+R_b\right)^N}{N!}
e^{-(R_f+R_b)},
\end{equation}
since the distribution of total event number is a Poisson process with
rate $R_f+R_b$.  Combining these yields the conditional probability of
the flags on the rates:
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{g_i},N | R_f, R_b\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N!} \exp\left[ - \left(R_f + R_b\right) \right].
\end{equation}

@@

Putting everything together, the posterior is
\begin{multline}
  \label{eq:posterior}
  p\left( \mathset{g_i}, R_f, R_b, \theta | d_{\rm to},N \right) =
  \frac{\alpha}{p(d_{\rm to},N)\,N!} \\ \times \left[ \prod_{\mathset{i | g_i =
        1}} R_f \hat{f}\left(x_i, \theta\right) \right] \left[
    \prod_{\mathset{i | g_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
    \right] \\ \times \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
When sampling the posterior, the first term, which is independent of
the parameters of interest, can be omitted and the equals sign
replaced by proportionality; however, we have kept this term
explicitly so that we can see the equivalence to ranking-statistic
ordered data. Once data have been observed, there is a unique loudness
ordering and time ordering of those events, and so there is a one to
one correspondence between a time-ordered posterior $p\left(
\mathset{g_i}, R_f, R_b, \theta | d_{\rm to},N \right)$ and the
corresponding statistic-ordered posterior $ p\left( \mathset{g_i},
R_f, R_b, \theta | d,N \right)$, which means $p\left( \mathset{g_i},
R_f, R_b, \theta | d,N \right) = p\left( \mathset{g_i}, R_f, R_b,
\theta | d_{\rm to},N \right)$. However, the evidence $p(d,N) =
N!\,p(d_{\rm to},N)$, since there are $N!$ ways in which $N$ events
with a given set of ranking statistics can be ordered in time.

The ranking-statistic ordered posterior can be computed directly by
assuming that the flags, $\mathset{g_i}$, are un-observed data and
treating the sets $\mathset{x_i | g_i = 1}$ and $\mathset{x_i | g_i =
  0}$ as samples from an inhomogeneous Poisson process.  For an
inhomogeneous Poisson process with rate function $r(y)$ (cumulative
rate $R(y)$), the likelihood of a set of samples $\mathset{y_i}$ is
given by
\begin{multline}
  p\left( \mathset{y_i} | r \right) \,{\rm d}^Ny_i = P\left(
  \textnormal{zero events below } y_1 \right) \\ \times P\left(
  \textnormal{one event between } y_1 \textnormal{ and } y_1 + {\rm
    d}y_1 \right) \\ \times P\left( \textnormal{zero events between }
  y_1+{\rm d}y_1 \textnormal{ and } y_2 \right) \ldots,
\end{multline}
so 
\begin{multline}
p\left(\mathset{y_i} | r \right) = \lim_{\delta y_i \to 0} \exp\left[ -
    R\left(y_1\right) \right] \left[ r\left( y_1 \right) +
    \order{\delta y_1} \right] \\ \times \exp\left[ - \left[ R\left( y_2
      \right) - R\left( y_1 + \delta y_1\right) \right] \right] \times
  \ldots \\ = \left[\prod_i r\left( y_i \right)\right] \exp\left[ -
    R\left( \infty \right) \right].
\end{multline}
Applying this once to the foreground samples, once to the background
samples and taking the product, we obtain $p(d,\mathset{g_i}, N | R_f,
R_b, \theta)$ and thus $p(\mathset{g_i}, R_f, R_b, \theta | d,N) =
p(d,\mathset{g_i}, N | R_f, R_b, \theta)\ p(R_f, R_b, \theta)\, /\,
p(d,N)$. With the identification $p(d,N) = N!\,p(d_{\rm to},N)$, as
justified above, we reproduce Eq.~\eqref{eq:posterior}.

We can marginalize the posterior over the flags, $g_i$, obtaining
\begin{multline}
  \label{eq:rate-shape-posterior}
  p\left( R_f, R_b, \theta | d, N \right) = \sum_{\mathset{g_i} \in
    \mathset{0,1}^N} p\left( \mathset{g_i}, R_f, R_b, \theta | d, N
  \right) \\ \propto \prod_{i} \left[ R_f \hat{f}\left(x_i, \theta\right)
    + R_b \hat{b}\left( x_i, \theta\right) \right] \\ \times \exp\left[-\left(
    R_f + R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
This expression is useful if we are only interested in rates and not
the probability that any particular event is foreground or background.
Unlike the full posterior (Eq.~\eqref{eq:posterior}),
Eq.~\eqref{eq:rate-shape-posterior} contains only continuous
parameters. We note that the terms that depend on the overall rate
parameters, $R_b$ or $R_f$, are of the form $R_b^{n-1/2} \exp(-R_b)$
and so marginalization over either $R_b$ or $R_f$ can be achieved
analytically using
\begin{equation}
I_n = \int_0^{\infty} x^{n-\frac{1}{2}} \, {\rm e}^{-x} {\rm d} x =
\frac{(2n-1)!!}{2^n} \sqrt{\pi}
\end{equation}
using the usual notation $(2n-1)!! \equiv (2n-1)(2n-3)\cdots1$.

@@

If $\tilde{b}(x_N,\theta) \ll \tilde{f}(x_N,\theta)$, we obtain the
result $(1- \hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be
understood as the statement that the rate of foreground events with
ranking statistic greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$,
is of order $1$, as expected. However, $\tilde{b}(x_N,\theta) = -{\rm
  d} [\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term may be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) \gg \tilde{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavored.
